---
title: "Onassis: Ontology Annotation and Semantic Similarity software"
author: "Eugenia Galeota"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{Onassis: Ontology Annotation and Semantic Similarity software}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignettePackage{Onassis}
  %\VignetteDepends{Onassis}
---
<style>
body {
text-align: justify}
</style>
```{r imports, echo=FALSE,eval=TRUE, message=FALSE, warning=FALSE}
library(Onassis)
library(DT)
library(gplots)
library(org.Hs.eg.db)
library(kableExtra)
```

# Introduction to OnASSis 
Public repositories of biological omics data contain thousands of experiments. While these resources are extrenely useful, those data are difficult to mine. The annotation the associated metadata with controlled vocabularies or ontology terms is essential to facilitate the retrieval of data of interest both programmatically or manually [@galeota2016ontology].
<b>OnASSiS</b> (Ontology Annotations and Semantic Similarity software) is a package aimed at matching metadata associated with biological experiments with concepts from ontologies, allowing the construction of semantically coherent omics datasets, possibly representing various data types from independent studies. The recognition of domain specific entities not only allows users to retrieve samples related to a given cell type or experimental condition, but also to unravel previously unanticipated relationships between experiments. Onassis applies Natural Language Processing tools to annotate sample's and experiments' descriptions, recognizing concepts from a multitude of biomedical ontologies and quantifying the similarities between pairs or groups of query studies. Moreover, it assists the semantically driven analysis of the corresponding omics data.
In particular the software includes modules to enable:

* the <b>retrieval</b> of samples’ metadata from repositories of large scale biologial data

* the <b>annotation</b> of these data with concepts belonging to Open Biomedical Ontologies (OBO) 

* the organization of selected samples in structured groups based on <b>semantic similarity</b> measures

* the <b>comparison</b> of biological signal (e.g. gene expression within different samples) based on the entities associated to each samnple and their relationship

Onassis relies on Conceptmapper, an Apache UIMA (Unstructured Information Management Architecture) dictionary lookup tool to retrieve dictionary terms in a given text. https://uima.apache.org/downloads/sandbox/ConceptMapperAnnotatorUserGuide/ConceptMapperAnnotatorUserGuide.html  
In particular, the ccp-nlp Conceptmapper wrapper, specific for the biomedical domain, implements a pipeline through which it is possible to retrieve concepts from OBO ontologies in any given text with different adjustable options [@Verspoor:2009b].

Onassis features can be easily accessed through a main class named Onassis, having as slots 'dictionary', 'entities', 'similarity' and 'scores'. In the following sections we first show details on the usage of the classes and methods that constitute the building blocks of typical metadata integration workflows. Next, we show how the Onassis class wraps all these functions for a simplified access and usage.

Onassis can handle any type of text as input, but is particularly well suited for the analysis of the metadata from Gene Expression Omnibus (GEO). Indeed it provides the possibility to associate concepts from any OBO ontology to GEO metadata retrieved using `r BiocStyle::Biocpkg("GEOmetadb")`. In general, any table or database (such as Sequence Read Archive (SRA) [@Zhu2013] or Cistrome [@Shengling2017]) containing textual descriptions that can be easily imported in R as a data frame can be used as input for Onassis. 
In addition to the ontological concepts, the recognition of gene/protein symbols or epigenetic modifications can be highly relevant, especially for experiments directed to specific factors or marks (such as ChIP-seq experiments).
The semantic similarity module uses different semantic similarity measures to determine the semantic similarity of concepts in a given ontology. This
module has been developed on the basis of the Java slib http://www.semantic-measures-library.org/sml.
The score module applies statistical tests to determine if biological signals from samples annotated with different concepts, from one or more ontologies, are significantly different.

# Installation

To run Onassis Java (>= 1.8) is needed. To install the package please run the following code
```{r installing_onassis, echo=TRUE, eval=FALSE}
source("https://bioconductor.org/biocLite.R")
biocLite("Onassis")
````

Onassis can be loaded with the following code
```{r load_onasssis, echo=TRUE, eval=TRUE}
library(Onassis)
```

Some of the optional functions, which will be described in the followign parts of the vignette, require additional libraries. These include:

* org.Hs.eg.db (Bioconductor)
* GenomicRanges (Bioconductor)
* gplots (CRAN)
* GEOmetadb (Bioconductor)
* SRAdb (Bioconductor)

# Retrieving public repositories metadata

One of the most straightforward ways to retrieve metadata of samples provided in GEO is through `r BiocStyle::Biocpkg("GEOmetadb")` package. In order to use GEOmetadb through Onassis, users should download the corresponding SQLite database file, by following the instructions provided in the vignette of that package. This step should be performed only once. Onassis provides functions to help GEO metadata retrieval without the need of explicitly making SQL queries to the database. While GEOmetadb can be accessed on any platform, another important database `r BiocStyle::Biocpkg("SRAdb")`, is not available for Windows users. In the following sections we show how to query GEOmetadb through Onassis, but we also provide an example on how to access SRAdb metadata. 

## Handling GEO (Gene Expression Omnibus) Metadata

First, it is necessary to obtain a connection to the SQLite database. If this was already installed, `connectToGEODB` returns a connection to the database given the full path to the SQLite database file. Alternatively, by setting `download` to TRUE the database is downloaded. The `getGEOmetadata` function can be used to retrieve the metadata related to specific GEO samples, taking as minimal parameters the connection to the database and one of the experiment types available. Optionally it is possible to specify the organism and the platform. For the purpose of this vignette, the queries illustrated below can be skipped, since the meth_metadata object was already stored within Onassis.

```{r connectTodb, echo=TRUE,eval=FALSE}

require('GEOmetadb')

## Running this function might take some time if the database (6.8GB) has to be downloaded.
geo_con <- connectToGEODB(download=TRUE)

#Showing the experiment types available in GEO
experiments <- experiment_types(geo_con)

#Showing the organism types available in GEO
species <- organism_types(geo_con)

#Retrieving the metadata associated to experiment type "Methylation profiling by high througput sequencing"
meth_metadata <- getGEOMetadata(geo_con, experiment_type='Methylation profiling by high throughput sequencing', organism = 'Homo sapiens')

#Retrieving Human gene expression metadata, knowing the GEO platform identifier, e.g. the Affymetrix Human Genome U133 Plus 2.0 Array
expression <- getGEOMetadata(geo_con, experiment_type='Expression profiling by array', gpl='GPL570')
```

Some of the experiment types available are the following:

```{r experimentTypesshow, echo=FALSE, eval=TRUE}
experiments <- readRDS(system.file('extdata', 'vignette_data', 'experiment_types.rds', package='Onassis'))

knitr::kable(as.data.frame(experiments[1:10]), col.names = c('experiments')) %>% kable_styling(bootstrap_options = c("striped"), position="center") %>%
  scroll_box(width = "300px", height = "200px")
```

Some of the organisms available are the following:

```{r speciesShow, echo=FALSE,eval=TRUE}
species <- readRDS(system.file('extdata', 'vignette_data', 'organisms.rds', package='Onassis'))
knitr::kable(as.data.frame(species[1:10]), col.names=c('species')) %>%
              kable_styling(bootstrap_options = c("striped"), position="center") %>%
  scroll_box(width = "300px", height = "200px")
```

As specified above, meth_metadata was previously saved and can be loaded from the Onassis package external data:

```{r loadgeoMetadata, echo=TRUE, eval=TRUE}
meth_metadata <- readRDS(system.file('extdata', 'vignette_data', 'GEOmethylation.rds', package='Onassis'))
```

```{r printmeta, echo=FALSE,eval=TRUE}

methylation_tmp <- meth_metadata
methylation_tmp$experiment_summary <- sapply(methylation_tmp$experiment_summary, function(x) substr(x, 1, 50))
knitr::kable(methylation_tmp[1:10,], 
              caption = 'GEOmetadb metadata for Methylation profiling by high throughput sequencing (only the first 10 entries are shown).') %>%
              kable_styling(bootstrap_options = c("striped"), position="center") %>%
  scroll_box(width = "80%", height = "300px")

```

## Handling SRA (Sequence Read Archive) Metadata

In this section we provide an example showing how it is possible to retrieve data from other sources such as SRA. This database is not directly supported by Onassis, since it is not available for Windows platforms. Hence, the code reported below is slightly more complicated, and exemplifies how to query the SRA database provided by the SRAdb package and store metadata of human ChIP-seq experiments within a data frame. Due to the size of the SRA database (36 GB), results of the query are available within Onassis as external data, and the example code below can be skipped.

```{r connectSRA, echo=TRUE,eval=FALSE}
# Optional download of SRAdb and connection to the corresponding sqlite file
require(SRAdb)
sqliteFileName <- '/pathto/SRAdb.sqlite'
sra_con <- dbConnect(SQLite(), sqliteFileName)

# Query for the ChIP-Seq experiments contained in GEO for human samples 
library_strategy <- 'ChIP-Seq' #ChIP-Seq data
library_source='GENOMIC' 
taxon_id=9606 #Human samples
center_name='GEO' #Data from GEO
 
# Query to the sample table 
samples_query <- paste0("select sample_accession, description, sample_attribute, sample_url_link from sample where taxon_id='", taxon_id, "' and sample_accession IS NOT NULL", " and center_name='", center_name, "'",  )

samples_df <- dbGetQuery(sra_con, samples_query)
samples <- unique(as.character(as.vector(samples_df[, 1])))

# Query to the experiment table
experiment_query <- paste0("select experiment_accession, center_name, title, sample_accession, sample_name, experiment_alias, library_strategy, library_layout, experiment_url_link, experiment_attribute from experiment where library_strategy='", 
                           library_strategy, "'" , " and library_source ='", library_source,
                           "' " )

experiment_df <- dbGetQuery(sra_con, experiment_query)

#Merging the columns from the sample and the experiment table
experiment_df <- merge(experiment_df, samples_df, by = "sample_accession")

# Replacing the field separators with white spaces
experiment_df$experiment_attribute <- sapply(experiment_df$experiment_attribute, 
                                             function(value) {
                                               gsub("||", "  ", value)
                                             })
experiment_df$sample_attribute <- sapply(experiment_df$sample_attribute, 
                                         function(value) {
                                           gsub("||", "  ", value)
                                         })
# Replacing the '_' character with white spaces
experiment_df$sample_name <- sapply(experiment_df$sample_name, 
                                    function(value) {
                                      gsub("_", " ", value)
                                    })
experiment_df$experiment_alias <- sapply(experiment_df$experiment_alias, 
                                         function(value) {
                                           gsub("_", " ", value)
                                         })
sra_chip_seq <- experiment_df
```

Alternatively, as described above, a previously saved table sra_chip_seq with 500 samples can be retrieved in Onassis:

```{r readCHIP, echo=TRUE, eval=TRUE}
sra_chip_seq <- readRDS(system.file('extdata', 'vignette_data', 'GEO_human_chip.rds',  package='Onassis'))
```

```{r printchromatinIP, echo=FALSE,eval=TRUE}

knitr::kable(head(sra_chip_seq, 10), rownames=FALSE, 
  caption = 'Metadata of ChIP-seeq human samples obtained from SRAdb (first 10 entries)') %>%
              kable_styling(bootstrap_options = c("striped"), position="center") %>%
  scroll_box(width = "80%", height = "300px")

```

# Annotating text with Ontology Concepts

The Onassis `EntityFinder` class has methods for annotating any text with dictionary terms. More specifically, Onassis can take advantage of the OBO dictionaries (http://www.obofoundry.org/).

## Data preparation

The findEntities method supports input text in the form of:

  * The path of a directory containing named documents. 
  * The path of a single file containing multiple documents. In this case each row contains the name/identifier of the document followed by a '|' separator and the text to annotate.

Alternatively, the annotateDF method supports input text in the form of a data frame. In this case each row represents a document, first column has to be the document identifier, and the remaining columns will be combined and contain the text to analyze. This option can be conveniently used with the metadata retrieved from `r BiocStyle::Biocpkg("GEOmetadb")` and `r BiocStyle::Biocpkg("SRAdb")`, possibly selecting a subset of the available columns.

## Creation of a Conceptmapper Dictionary

Conceptmapper dictionaries are XML files with a set of entries specified by the xml tag ```<token>```  with a canonical name (the name of the entry) and one or more variants (synonyms).

The constructor `CMdictionary` creates an instance of the class `CMdictionary`.

  * If an XML file containing the Conceptmapper dictionary is already available, it can be uploaded into Onassis indicating its path and setting the `dictType` option to "CMDICT".
  * If the dictionary has to be built from an OBO ontology (OBO or OWL formats are supported), the path or URL to the corresponding file has to be provided and dictType has to be set to "OBO". The synonymType argument can be set to EXACT_ONLY or ALL to consider only canonical concept names or also to include any synonym. The resulting XML file is written in the indicated outputdir. 
  * Additionally, to facilitate the named entity recognition of specific targets, such in the case of ChIP-seq experiments, these can be included within a specific dictionary, and dictType has to be set to ENTREZ. If a specific Org.xx.eg.db Bioconductor library is installed and loaded, it can be indicated in the inputFileOrDb parameter as a character string, and gene names will be derived from it. Instead, if inputFileOrDb is empty and a specific species is indicated in the taxID parameter, gene names will be derived from the corresponding gene_info.gz file downloaded from NCBI (300MB). Finally, if dictType is set to  TARGET, known histone post-translational modifications and epigenetic marks are also included.

```{r createSampleAndTargetDict, echo=TRUE,eval=TRUE, message=FALSE}
# If a Conceptmapper dictionary is already available the dictType CMDICT can be specified and the corresponding file loaded
sample_dict <- CMdictionary(inputFileOrDb=system.file('extdata', 'cmDict-sample.cs.xml', package = 'Onassis'), dictType = 'CMDICT')

#Creation of a dictionary from the file sample.cs.obo available in OnassisJavaLibs
obo <- system.file('extdata', 'sample.cs.obo', package='OnassisJavaLibs')

sample_dict <- CMdictionary(inputFileOrDb=obo, outputDir=getwd(), synonymType='ALL')

# Creation of a dictionary for human genes/proteins. This requires org.Hs.eg.db to be installed
require(org.Hs.eg.db)
targets <- CMdictionary(dictType='TARGET', inputFileOrDb = 'org.Hs.eg.db')
```
 
 The following XML markup code illustrates a sample of the Conceptmapper dictionary corresponding to the Brenda tissue ontology.

```xml 
   <?xml version="1.0" encoding="UTF-8" ?>
   <synonym>
      <token id="http://purl.obolibrary.org/obo/BTO_0005205" canonical="cerebral artery">
        <variant base="cerebral artery"/>
      </token>
      <token id="http://purl.obolibrary.org/obo/BTO_0002179" canonical="184A1N4 cell">
        <variant base="184A1N4 cell"/>
        <variant base="A1N4 cell"/>
      </token>
      <token id="http://purl.obolibrary.org/obo/BTO_0003871" canonical="uterine endometrial cancer cell">
        <variant base="uterine endometrial cancer cell"/>
        <variant base="endometrial cancer cell"/>
        <variant base="uterine endometrial carcinoma cell"/>
        <variant base="endometrial carcinoma cell"/>
      </token>
  </synonym>
```

## Setting the options for the annotator

Conceptmapper includes 7 different options controlling the annotation step. These are documented in detail in the documentation of the CMoptions function. They can be listed through the `listCMOptions` function. The `CMoptions` constructor instantiates an object of class CMoptions with the different parameters that will be required for the subsequent step of annotation. We also provided getter and setter methods for each of the 7 parameters.  

```{r settingOptions, echo=TRUE,eval=TRUE}

#Creating a CMoptions object and showing hte default parameters 
opts <- CMoptions()  
show(opts)
```

To obtain the list of all the possible combinations:

```{r listCombinations, echo=TRUE, eval=TRUE}
combinations <- listCMOptions()
```

To create a CMoptions object having has SynonymType 'EXACT_ONLY', that considers only exact synonyms, rather than 'ALL' other types included in OBO (RELATED, NARROW, BROAD) 

```{r setsynonymtype, echo=TRUE, eval=TRUE}
myopts <- CMoptions(SynonymType = 'EXACT_ONLY')
myopts
```

To change a given parameter, for example to use a search strategy based on the Longest match of not-necessarily contiguous tokens where overlapping matches are allowed:

```{r changeparameter, echo=TRUE, eval=TRUE}
#Changing the SearchStrategy parameter
SearchStrategy(myopts) <- 'SKIP_ANY_MATCH_ALLOW_OVERLAP'
myopts
```


## Running the entity finder

The class `EntityFinder` defines a type system and runs the Conceptmapper pipeline. It can find concepts of any OBO ontology in a given text. The `findEntities` and `annotateDF` methods accept text within files or data.frame, respectively, as described in Section 4.1.
The function `EntityFinder` automatically adapts to the provided input type, creates an instance of the `EntityFinder` class to initialize the type system and runs the pipeline with the provided options and dictionary. 
For example, to annotate the metadata derived from ChIP-seq experiments obtained from SRA with tissue and cell type concepts belonging to the sample ontology available in Onassis and containing tissues and cell names, the following code can be used:

```{r EntityFinder, echo=TRUE, eval=TRUE, results='hide', message=FALSE, warning=FALSE }
chipseq_dict_annot <- EntityFinder(sra_chip_seq[1:50, c('sample_accession', 'title', 'experiment_attribute', 'sample_attribute', 'description')], dictionary=sample_dict, options=myopts)
```

The resulting data.frame contains, for each row, a match to the provided dictionary for a specific document/sample (indicated in the first column). The annotation is reported with the id of the concept (term_id), its canonical name (term name), its URL in the obo format, and the matching sentence of the document.

```{r showchipresults, echo=FALSE, eval=TRUE, message=FALSE}

knitr::kable(head(chipseq_dict_annot, 20), rownames=FALSE, 
  caption = 'Annotations of the methylation profiling by high througput sequencing metadata obtained from GEO with BRENDA ontology concepts') %>% kable_styling() %>%
  scroll_box(width = "80%", height = "400px")
```

\r
\r
The function `EntityFinder` can also be used to identify the targeted entity of each ChIP-seq experiment, by retrieving gene names and histone types or modifications in the ChIP-seq metadata.   

```{r annotateGenes, echo=TRUE, eval=TRUE, results='hide', message=FALSE, warning=FALSE}
#Finding the TARGET entities
target_entities <- EntityFinder(input=sra_chip_seq[1:50, c('sample_accession', 'title', 'experiment_attribute', 'sample_attribute', 'description')], options = myopts, dictionary=targets) 
```

```{r printKable, echo=FALSE, eval=TRUE}
knitr::kable(head(target_entities, 20), 
  caption = 'Annotations of ChIP-seq test metadata obtained from SRAdb and stored into files with the TARGETs (genes and histone variants)') %>% kable_styling(bootstrap_options = c("striped"), position="center") %>%
  scroll_box(width = "80%", height = "400px")
```

# Semantic similarity 

Once a set of samples is annotated, i.e. associated to a set of ontology concepts, Onassis allows the quantification of the similarity among these samples based on the semantic similarity between the corresponding concepts. `Similarity` is an Onassis class applying methods of the Java library slib \url{http://www.semantic-measures-library.org/sml/} [@harispe2014semantic], which builds a semantic graph starting from OBO ontology concepts and their hierarchical relationships.
The following methods are available and are automatically chosen depending on the settings of the `Similarity` function. The `sim` and `groupsim` methods allow the computation of semantic similarity between single terms (pairwise measures) and between group of terms (groupwise measures), respectively. Pairwise measures can be edge based, if they rely only on the structure of the ontology, or information-content based if they also consider the information that each term in the ontology carries. Rather, groupwise measures can be indirect, if they compute the pairwise similarity between each couple of terms, or direct if they consider each set of concepts as a whole.
The `samplesim` method allows to determine the semantic similarity between two documents, each possibly associated to multiple concepts, using groupwise measures. Finally, the `multisim` method allows to determine the semantic similarity between documents annotated with two or more ontologies: first `samplesim` is run for each ontology, then a user defined function can be used to aggregate the resulting semantic similarities for each pair of documents.  

The function `listSimilarities` shows all the measures supported by Onassis. For details about the measures run `{?Similarity}`.

```{r similarity, echo=TRUE, eval=TRUE, message=FALSE}
#Instantiating the Similarity
similarities <- listSimilarities()
```

The following example shows the pairwise similarities between sample cell concepts obtained annotating the ChIP-seq metadata. The <b>lin</b> similarity measure is used by default, which relies on a ratio between the Information content (IC) of the Most Specific Common Ancestor of the compared concepts and the sum of their IC based on the information content of the most informative common ancestor of the considered concepts. In particular, the <b> seco </b> information content is used by default, which determines the specificity of each concept based on the number of concepts it subsumes. 

```{r computing measures, echo=TRUE, eval=TRUE, message=FALSE}

found_terms <- as.character(unique(chipseq_dict_annot$term_url))
n <- length(found_terms)

ontologyfile <- obo

pairwise_results <- data.frame( matrix(0, nrow=n * (n-1) / 2, ncol=3))
colnames(pairwise_results) <- c('term1', 'term2', 'value')  
  
#  term1 = character(0), term2= character(0), value = double(0L))
row_count <- 0 
for(i in 1:(n-1)){
  j = i + 1 
  for(k in j:n){
    two_term_similarity <- Similarity(ontologyfile, as.character(found_terms[i]), as.character(found_terms[k]) )
    row_count <- row_count + 1 
    pairwise_results[row_count, ] <- c(as.character(found_terms[i]), as.character(found_terms[k]), two_term_similarity)
  }
}
pairwise_results <- merge(pairwise_results, chipseq_dict_annot[, c('term_url', 'term_name')], by.x='term2', by.y='term_url')
colnames(pairwise_results)[length(colnames(pairwise_results))] <- 'term2_name'
pairwise_results <- merge(pairwise_results, chipseq_dict_annot[, c('term_url', 'term_name')], by.x='term1', by.y='term_url')
colnames(pairwise_results)[length(colnames(pairwise_results))] <- 'term1_name'
pairwise_results <- unique(pairwise_results)
pairwise_results <- pairwise_results[, c('term1', 'term1_name', 'term2', 'term2_name', "value")]
```



```{r showSim, echo=FALSE, eval=TRUE}
knitr::kable(pairwise_results,  
  caption = 'Pairwise similarities of cell line terms annotating the ChIP-seq metadata') %>% kable_styling(bootstrap_options = c("striped"), position="center") %>%
  scroll_box(width = "80%", height = "400px")
```

Noteworthy, the terms 'B-cell' and 'lymphocyte' are closer (similarity 0.83) than 'B cell' and 'epithelial cell' (similarity 0.26).
It is also possible to compute the semantic similarity between two groups of terms. 
For example, to determine a value of similarity for the combination of 'non-terminally differentiated cell', 'epithelial cell' and the combination of 'lymphocyte' and 'B cell' we can use the <b>ui</b> measure (set as default measure in Onassis), a groupwise direct measure combining the intersection and the union of the set of ancestors of the two groups of concepts.  

```{r groupwise_measures, echo=TRUE, eval=TRUE, message=FALSE}

Similarity(obo, c('http://purl.obolibrary.org/obo/CL_0000055', 'http://purl.obolibrary.org/obo/CL_0000066'), c('http://purl.obolibrary.org/obo/CL_0000542' ,'http://purl.obolibrary.org/obo/CL_0000236'))
```

The similarity between the two groups of terms is quiet low (1.83), while replacing    


```{r groupwise_measures, echo=TRUE, eval=TRUE, message=FALSE}

Similarity(obo, c('http://purl.obolibrary.org/obo/CL_0000055', 'http://purl.obolibrary.org/obo/CL_0000236' ), c('http://purl.obolibrary.org/obo/CL_0000542' , 'http://purl.obolibrary.org/obo/CL_0000066'))

```



Lastly, the pariwise semantic similarity between ChIP-seq samples is illustrated.

```{r samples_similarity, echo=TRUE, eval=TRUE, message=FALSE}

annotated_samples <- as.character(as.vector(unique(chipseq_dict_annot$sample_id)))
n <- length(annotated_samples)


samples_results <- data.frame(sample1 = character(0), sample2= character(0), value = double(0L))
samples_results <- matrix(0, nrow=n, ncol=n)
rownames(samples_results) <- colnames(samples_results) <- annotated_samples
for(i in 1:(n-1)){
  sample1 <- as.character(annotated_samples[i])
  j = i + 1 
  for(k in j:n){
    sample2 <- as.character(annotated_samples[k])
    two_samples_similarity <- Similarity(ontologyfile, sample1, sample2, chipseq_dict_annot)
    samples_results[i, k] <- samples_results[k, i] <- two_samples_similarity
  }
}
diag(samples_results) <- 1
heatmap.2(samples_results, density.info = "none", trace="none", main='Semantic similarity of annotated samples', margins=c(5,5))
```

# Onassis class

The class Onassis was built to wrap the main functionalities of the package in a single class.
It consists of 4 slots:

* dictionary: stores the source dictionary used to find entities.

* entities: a table containing the annotations of documents (samples). The list of unique concepts belonging to the dictionary and found in the metadata representing a given sample is defined as a that sample semantic set

* similarity: a matrix of the similarities between the unique semantic sets identified in the entities table

* scores: a dataset of quantitative measurements (e.g. gene expression) associated to the same samples annotated in the entities slot.
 
In this section we illustrate the use of the Onassis class to annotate the previously retrieved metadata. The method `annotate` takes as input a data frame of metadata to annotate, the type of dictionary and the path of an ontology file and returns an instance of class Onassis.
The input data frame should have unique identifiers in the first column (sample identifiers or generic document identifiers) and for each row one or more columns containing the metadata related to the identifier. 


```{r onassis_class_usage, echo=TRUE, eval=TRUE, results='hide', message=FALSE, warning=FALSE }
onassis_annotations <- annotate(sra_chip_seq, 'OBO',obo )
```

To slot `entities` of an Onassis object reports for each sample the unique list of concepts found in the corresponding metadata. Specifically, term ids, term urls and term names are reported, and multiple entries per sample are comma separated. We usually refer to these unique lists as semantic sets.

To retrieve the semantic sets in an object of class Onassis we provided the accessor method `entities`

```{r show_onassis_annotations, echo=TRUE, eval=TRUE}
onassis_entities <- entities(onassis_annotations) 
```

```{r showing_entities, echo=FALSE, eval=TRUE}
knitr::kable(
onassis_entities[sample(nrow(onassis_entities), 10),], 
  caption = 'Semantic sets of ontology concepts (entities) associated to each sample, sotred in the entities slot of the Onassis object') %>% kable_styling(bootstrap_options = c("striped"), position="center") %>%
  scroll_box(width = "80%", height = "400px")
```

The `filterconcepts` method can be used to filter out unwanted annotations, for example terms that we consider redundant or too generic. The method modifies the entities slot of the Onassis object and returns a new Onassis object with filtered semantic sets.

```{r term_filtering, echo=TRUE, eval=TRUE}
filtered_onassis <- filterconcepts(onassis_annotations, c('cell'))
```

```{r showing_filt_entities, echo=FALSE, eval=TRUE}
knitr::kable(entities(filtered_onassis), 
  caption = 'Entities in filtered Onassis object') %>% kable_styling(bootstrap_options = c("striped"), position="center") %>%
  scroll_box(width = "80%", height = "400px")
```

The method `sim` populates the similarity slot within an Onassis object. Specifically, it generates a matrix containing semantic similarities between the semantic sets for each pair of samples annotated in the entities slot. 

```{r similarity_of_samples, echo=TRUE, eval=TRUE}
filtered_onassis <- sim(filtered_onassis)
```

The matrix of similarities can be accessed using the method `simil(filtered_onassis)`.

Semantic sets with semantic similarities above a given threshold can be combined using the method `collapse`. This method, based on hierarchical clustering, unifies the similar semantic sets by concatenating their unique concepts. Term names and term urls in the `entities` slot will be updated accordingly. For each concept, the number of samples associated is also reported in bracket squares, while the total number of samples associated to a given semantic set is indicated in parentheses.
After the collapse, the similarity matrix in the `similarity` slot is consequently updated with the similarities of the new semantic sets.

```{r collapsing_similarities, echo=TRUE, eval=TRUE, message=FALSE, results='hide', fig.width=6, fig.height=6}
collapsed_onassis <- Onassis::collapse(filtered_onassis, 0.8)
head(entities(collapsed_onassis))

heatmap.2(simil(collapsed_onassis), margins=c(15,15), cexRow = 1, cexCol = 1)
```

In typical integrative analyses scenarios, users could be interested in annotating concepts from different domains of interest. In this case, one possibility could be building a tailored application ontology including the concepts from different ontologies and their relationships. However this is complicated, since it requires to match and integrate the relationships from one ontology to the other. Rather, Onassis allows integrating two ontologies by repeating the annotation process with another ontology, while keeping separate the semantic sets from the two ontologies. In the following example ChIP-seq samples will be annotated with information about disease conditions. For this particular semantic type, Onassis provides also a boolean variable `disease` that can be set to TRUE to recognize samples metadata explicitly annotated as `Healthy` conditions, to be differentiated from metadata where disease terms are simply lacking.

```{r creating_disease_annotations, echo=TRUE, eval=TRUE, message=FALSE, results='hide'}
obo2 <- system.file('extdata', 'sample.do.obo', package='OnassisJavaLibs')
disease_annotations <- annotate(sra_chip_seq, 'OBO',obo2, disease=TRUE )
```

The method `mergeonassis` can be used to combine two Onassis objects in which the same set of samples was annotated with two different ontologies. This is useful to perform a nested analysis driven by the annotation provided by the two ontologies. The first object associates the samples metadata to semantic sets from a primary domain of interest. For each semantic set, the associated samples will be further separated based on the semantic sets belonging to a secondary domain. For example, users could be interested in comparing different diseases (secondary domain) within each cell line (primary domain), for a set of cell lines. 

```{r merging_onassis_entities, echo=TRUE, eval=TRUE, message=FALSE}
cell_disease_onassis <- mergeonassis(collapsed_onassis, disease_annotations)
```

The method `compare` exploits the nested object created with `mergeonassis` to analyse the actual omics data. 
To this end, a score matrix containing measurements for arbitrary genomic units in the annotated samples is necessary. For example, genomics units can be genes, for expression analyses, or genomic regions (such as promoters) for enrichment analyses (such as ChIP-seq experiments).
We suggest to take advantage of data repositories where omics data were reanalysed with standardazied analysis pipelines. This would minimize confunding issues related to the differences due to alterantive analysis and normalization procedures. For example CistromeDB [@Shengling2017], provides ChIP-seq data from GEO re-analyzed with a standardized pipeline. We decided to use peak files related to the previously annotated samples from Cistrome to show an example of how `compare` works. 
Cistrome samples can be identified by GEO sample GSMs identifiers, which were matched with those reported in the SRA ChIP-seq metadata table. We precomputed a score matrix having as rows human promoter regions in chromosome 1 and as columns the sample identifiers. Each entry of the matrix contains the peak score value reported by Cistrome if there was a peak for a given promoter in a given sample. The score table can be retrieved from the vignette_data

```{r loading_score_matrix, echo=TRUE, eval=TRUE}

score_matrix <- readRDS(system.file('extdata', 'vignette_data', 'score_matrix.rds', package='Onassis'))

```

The method `compare` can be used with any test function to compare the scores for each genomic unit, based on the semantic sets associated to the corresponding samples. For example we created a function that summarized, for each promoter region, the peak scores from different samples in one semantic set with the mean peak score and then calls the `t.test` function. 

To apply the t.test for different diseases within cell line semantic sets the following code can be used. When by is set to 'col' the test function will be applied to columns, otherwise to genomic units belonging to different couples of semantic sets. Further details and examples about `compare` are available in the help page of the method. The method returns, in case of annotations with multiple ontologies a named list with the semantic sets of the first level ontology. Each entry of the list contains a matrix with the results of the test for each couple of second level semantic sets.  

```{r score_comparison, echo=T, eval=T}

personal_t <- function(x, y){
		if(is.matrix(x))
			x <- apply(x, 1, mean)
		if(is.matrix(y))
			y <- apply(y, 1, mean)
		do.call(match.fun('t.test'), list(x, y))
}

disease_comparisons <- compare(cell_disease_onassis, score_matrix=as.matrix(score_matrix), by='col', fun_name='personal_t')
head(disease_comparisons)[[1]][1,2]
```

# Session Info


Here is the output of sessionInfo() on the system on which this document was compiled through kintr:

```{r sessionInfo(), echo=FALSE, eval=TRUE}
sessionInfo()
```
# References
